---
title: "Historical Data Gathering and Cleaning for Machine Learning"
author: "Odin Lee"
format: html
editor: visual
---

# Load Libraries

```{r}

library(baseballr)
library(dplyr)
library(stringi)
library(purrr)

```

# Create Functions

### Function: Convert Baseball Style Innings Notation to Decimal Format

```{r}

# Example: "7.2" represents 7 innings and 2 outs, which is numerically 7 2/3 innings.

# Function to convert baseball-style innings pitched notation to a numerical value
convertInnings <- function(InningsPitched) {
  IP <- strsplit(as.character(InningsPitched), "\\.")[[1]]  # Split input on decimal point
  
  if (length(IP) == 2) {  # If there is a decimal portion
    wholeNumber <- as.numeric(IP[1])  # Extract whole innings
    fractionalNumber <- as.numeric(IP[2])  # Extract outs portion
    
    # Convert the outs portion to its fractional equivalent
    if (fractionalNumber == 1) {
      fractionalNumber <- 1/3  # 1 out = 1/3 of an inning
    } else if (fractionalNumber == 2) {
      fractionalNumber <- 2/3  # 2 outs = 2/3 of an inning
    }
    
    return(wholeNumber + fractionalNumber)  # Return total innings as a numeric value
  } else {
    return(as.numeric(InningsPitched))  # If no decimal, return as a numeric value
  }
}

```

### Function: Standardize Outfield Positions (LF, RF, and CF –\> OF)

```{r}

# Converts "LF" (Left Field), "RF" (Right Field), and "CF" (Center Field) to "OF" (Outfield)
positionToOF <- function(data) {
  data <- data |>
    mutate(position_abbreviation = ifelse(
      position_abbreviation %in% c("LF", "RF", "CF"),  # Check if position LF, RF, or CF
      "OF",  # Replace with "OF" for standardization
      position_abbreviation  # Keep other positions unchanged
    ))
  return(data)
}

```

### Function: Standardize Pitcher Position (P/TWP —\> SP)

```{r}

# Converts all "P" (Pitcher) and "TWP" (Two Way Player) designations to "SP" (Starting Pitcher)
pitchersToSP <- function(data) {
  data <- data |>
    mutate(position_abbreviation = ifelse(
      position_abbreviation %in% c("P", "TWP"),  # Check if position is listed as "P" or "TWP
      "SP",  # Replace with "SP" for standardization
      position_abbreviation  # Keep other positions unchanged
    ))
  return(data)
}

```

### Function: Reclassify Starters as Relievers (SP —\> RP)

```{r}

# Changes "SP" (Starting Pitcher) to "RP" (Relief Pitcher) if:
#     - Innings per game (I/G) is less than 4, OR
#     - Percentage of games started (GS/G) is less than 0.5

spToRP <- function(data) {
  data <- data |>
    mutate(position_abbreviation = ifelse(
      position_abbreviation == "SP" &  # Check if player is currently labeled as "SP"
      (innings_per_game < 4 | percent_games_started < 0.5),  # Check conditions
      "RP",  # Change to "RP" (Relief Pitcher)
      position_abbreviation  # Keep other positions unchanged
    ))
  return(data)
}

```

# Data Cleaning

#### Select Year

```{r}

# for multiple years do 2023:2025 this includes all three seasons
# singular season also works ie 2026

########## SELECT YEAR HERE ########## 

year = 2023:2025

########## SELECT YEAR HERE ##########

```

### Retrieve MLB Player Statistics

```{r}

# Fetch all hitter statistics
Hitters <- map_dfr(year, function(yr) {
  mlb_stats(
    stat_type = 'season',
    stat_group = 'hitting',
    season = yr,
    player_pool = 'all',
    limit = 1000
  )
})

# Fetch all pitcher statistics
Pitchers <- map_dfr(year, function(yr) {
  mlb_stats(
    stat_type = 'season',
    stat_group = 'pitching',
    season = yr,
    player_pool = 'all',
    limit = 1000
  )
})

```

### Change X Hitter Position

```{r}

# Change any position value == X to the players last known position
Hitters <- Hitters |>
  group_by(`player_full_name`) |>
  arrange(season) |>
  mutate(position_abbreviation = ifelse(position_abbreviation == "X", lag(position_abbreviation), position_abbreviation)) |>
  ungroup()

```

### Calculate Pitching Metrics

```{r}

# Computes innings pitched per game and percentage of games started for each pitcher
Pitchers <- Pitchers |> 
  mutate(innings_pitched = sapply(innings_pitched, convertInnings)) |>
  mutate(
    innings_per_game = as.numeric(innings_pitched) / games_played,  # Calculate IP per game
    percent_games_started = games_started / games_played  # Compute percentage of games started
  )

```

### Apply Functions

```{r}

# Standardize outfield positions in the hitters dataset (LF, RF, CF → OF)
Hitters <- positionToOF(Hitters)

# Convert all general pitcher positions ("P") to starting pitchers ("SP")
Pitchers <- pitchersToSP(Pitchers)

# Reclassify some starting pitchers ("SP") as relief pitchers ("RP") based on innings/game and start percentage
Pitchers <- spToRP(Pitchers)

```

### Select Relevant Columns for Fantasy Baseball Scoring

```{r}

# Simplify the Hitters dataset by selecting only necessary columns for FPts calculation
Hitters <- Hitters |>
  select(
    "Year" = "season",
    "Player Name" = "player_full_name", 
    "Position" = "position_abbreviation", 
    "Team" = "team_name",
    "Hits (H)" = "hits", 
    "Runs Scored (R)" = "runs", 
    "Total Bases (TB)" = "total_bases",
    "Runs Batted In (RBI)" = "rbi", 
    "Walks (BB)" = "base_on_balls", 
    "Intentional Walks (IBB)" = "intentional_walks", 
    "Strikeouts (K)" = "strike_outs",
    "Stolen Bases (SB)" = "stolen_bases"
  )

# Simplify the Pitchers dataset by selecting only necessary columns for FPts calculation
Pitchers <- Pitchers |>
  select(
    "Year" = "season",
    "Player Name" = "player_full_name", 
    "Position" = "position_abbreviation", 
    "Team" = "team_name",
    "Innings Pitched (IP)" = "innings_pitched", 
    "Hits Allowed (H)" = "hits", 
    "Earned Runs (ER)" = "earned_runs", 
    "Walks Issued (BB)" = "base_on_balls", 
    "Strikeouts (K)" = "strike_outs", 
    "Saves (SV)" = "saves", 
    "Blown Saves (BS)" = "blown_saves", 
    "Holds (HD)" = "holds"
  )

```

### Calculate Fantasy Points

```{r}

# Calculate fantasy points for hitters based on league scoring rules
Hitters <- Hitters |> 
  mutate(Team = ifelse(Team == "Oakland Athletics", "Athletics", Team)) |>
  mutate(`Fantasy Points (FPTS)` = 
           (`Hits (H)` * 1) + 
           (`Runs Scored (R)` * 1) + 
           (`Total Bases (TB)` * 1) + 
           (`Runs Batted In (RBI)` * 1) + 
           (`Walks (BB)` * 2) + 
           (`Intentional Walks (IBB)` * 1) + 
           (`Strikeouts (K)` * -1) + 
           (`Stolen Bases (SB)` * 2)
  ) |> 
  arrange(desc(`Fantasy Points (FPTS)`)) # Rank hitters from highest to lowest fantasy points

# Calculate fantasy points for pitchers based on league scoring rules
Pitchers <- Pitchers |> 
  mutate(Team = ifelse(Team == "Oakland Athletics", "Athletics", Team)) |>
  mutate(`Fantasy Points (FPTS)` = 
           (as.numeric(`Innings Pitched (IP)`) * 4.5) +
           (`Hits Allowed (H)` * -1) + 
           (`Earned Runs (ER)` * -3) + 
           (`Walks Issued (BB)` * -1) + 
           (`Strikeouts (K)` * 1) + 
           (`Saves (SV)` * 6) +
           (`Blown Saves (BS)` * -3) +
           (`Holds (HD)` * 5)
  ) |> 
  arrange(desc(`Fantasy Points (FPTS)`)) # Rank pitchers from highest to lowest fantasy points

```

### Select Relevant Rows

```{r}

#
# Numbers for the following filters can be found in file TeamScoresAnalysis.xlsx
# Locate the amount of hits or innings pitched for the player in the newest year 
# at the replacement level x1.20
#

# Filter out rows in Hitters that could cause skewing in ML model
Hitters <- Hitters |>
  # Filter out C with fewer than 313 FPTS. 
  # (313 = The number of FPTS by the average amount of C on a fantasy roster x1.2)
  filter(Position != "C" | (Position == "C" & `Fantasy Points (FPTS)` >= 313)) |>
  # Filter out 1B with fewer than 246 FPTS.  
  # (246 = The number of FPTS by the average amount of 1B on a fantasy roster x1.2)
  filter(Position != "1B" | (Position == "1B" & `Fantasy Points (FPTS)` >= 246)) |>
  # Filter out 2B with fewer than 269 FPTS.  
  # (269 = The number of FPTS by the average amount of 2B on a fantasy roster x1.2)
  filter(Position != "2B" | (Position == "2B" & `Fantasy Points (FPTS)` >= 269)) |>
  # Filter out 3B with fewer than 317 FPTS.  
  # (317 = The number of FPTS by the average amount of 3B on a fantasy roster x1.2)
  filter(Position != "3B" | (Position == "3B" & `Fantasy Points (FPTS)` >= 317)) |>
  # Filter out SS with fewer than 287 FPTS.  
  # (287 = The number of FPTS by the average amount of SS on a fantasy roster x1.2)
  filter(Position != "SS" | (Position == "SS" & `Fantasy Points (FPTS)` >= 287)) |>
  # Filter out OF with fewer than 336 FPTS.  
  # (336 = The number of FPTS by the average amount of OF on a fantasy roster x1.2)
  filter(Position != "OF" | (Position == "OF" & `Fantasy Points (FPTS)` >= 336)) |>
  # Filter out DH with fewer than 384 FPTS.  
  # (384 = The number of FPTS by the average amount of DH on a fantasy roster x1.2)
  filter(Position != "DH" | (Position == "DH" & `Fantasy Points (FPTS)` >= 384)) |>
  # Filter out any stragglers
  filter(`Fantasy Points (FPTS)` >= 100)

# Filter out rows in Pitchers that could cause skewing in ML model
Pitchers <- Pitchers |>
  # Filter out SP with fewer than 244.5 FPTS.  
  # (244.5 = The number of FPTS by the average amount of SP on a fantasy roster x1.2)
  filter(Position != "SP" | (Position == "SP" & `Fantasy Points (FPTS)` >= 244.5)) |>
  # Filter out RP with fewer than 362.5 FPTS.  
  # (362.5 = The number of FPTS by the average amount of RP on a fantasy roster x1.2)
  filter(Position != "RP" | (Position == "RP" & `Fantasy Points (FPTS)` >= 362.5)) |>
  # Filter out any stragglers
  filter(`Fantasy Points (FPTS)` >= 100)

```

#### Remove Letter Accents from Names

```{r}

Hitters$`Player Name` <- stri_trans_general(Hitters$`Player Name`, "Latin-ASCII")

Pitchers$`Player Name` <- stri_trans_general(Pitchers$`Player Name`, "Latin-ASCII")

```

#### Separate RP and SP

```{r}

# Create Starters with only SP
Starters <- Pitchers |>
  filter(Position == 'SP')

# Create Relievers with only RP
Relievers <- Pitchers |>
  filter(Position == 'RP')

```

# Export to CSV File

```{r}

# Export CSV files without row names
write.csv(Hitters, "HittersHistorical.csv", row.names = FALSE)
write.csv(Pitchers, "PitchersHistorical.csv", row.names = FALSE)
write.csv(Starters, "StartersHistorical.csv", row.names = FALSE)
write.csv(Relievers, "RelieversHistorical.csv", row.names = FALSE)

```
